Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2019-12-30 manuscript/appendix.Rmd
appendix changes

  frag_stats <- rbind(landscape_stats(historical),
                 landscape_stats(contemporary))
```

@@

>>>>

```

@@

```{r echo=FALSE, out.height='75%', fig.cap="CNN classifier results as run on a recent (2012) Geo-Eye panchromatic image. Results are compared with the Landsat based ."}
knitr::include_graphics('./figures/visual_comparison_classifiers.png')
```

```{r echo = FALSE, eval = TRUE}
# this data will not be available in the github repo due to missing geo-eye data

# read in cnn forest mask
m <- raster("../data/geo-eye/geo-eye_forest_mask_cnn.tif")

# hansen
hans <- raster("../data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")

# read in Hansen map (get all forest loss before 2013)
hans <- raster("../data/Hansen_et_al/Hansen_GFC_lossyear_yangambi.tif")
hans <- crop(hans, extent(m))
loss <- (hans <= 12) + (hans == 0)
loss <- (loss > 1)

ct <- crosstab(m, loss)
ct <- round(ct/sum(ct),2)
rownames(ct) <- c("non-forest","forest")

ct %>%
   kable(
    caption="Contingency table between the two forest / non-forest maps generated from a Geo-Eye 1 pan-chromatic image using a Convolutional Neural Network (CNN) generated and the Global Forest Cover map (Hansen et al. 2013).",
    booktabs = TRUE,
    col.names =  c("non-forest","forest")) %>%
  add_header_above(c("CNN" = 1, "Global Forest Cover" = 2)) %>%
  kable_styling(latex_options = c("hold_position"))

```

### 2019-12-30 manuscript/appendix.tex
appendix changes

\documentclass[landscape]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx}
% grffile has become a legacy package: https://ctan.org/pkg/grffile
\IfFileExists{grffile.sty}{%
\usepackage{grffile}
}{}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{figure}
\includegraphics[width=82.67in,height=0.75\textheight]{./figures/flight_paths_meta_data} \caption{Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-4}
\end{figure}

@@

>>>>

\documentclass[landscape]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{figure}
\includegraphics[height=0.75\textheight]{./figures/flight_paths_meta_data} \caption{Overview of the complete flight plan meta-data as stored in the archives at the Africa Museum.}\label{fig:unnamed-chunk-4}
\end{figure}

@@

\begin{figure}
\includegraphics[width=31.25in,height=0.75\textheight]{./figures/visual_comparison_classifiers} \caption{CNN classifier results as run on a recent (2012) Geo-Eye panchromatic image. Results are compared with the Landsat based .}\label{fig:unnamed-chunk-9}
\end{figure}

\caption{\label{tab:unnamed-chunk-10}Contingency table between the two forest / non-forest maps generated from a Geo-Eye 1 pan-chromatic image using a Convolutional Neural Network (CNN) generated and the Global Forest Cover map (Hansen et al. 2013).}
\centering
\begin{tabular}[t]{lrr}
\toprule
\multicolumn{1}{c}{CNN} & \multicolumn{2}{c}{Global Forest Cover} \\
\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-3}
  & non-forest & forest\\
\midrule
non-forest & 0.10 & 0.09\\
forest & 0.04 & 0.77\\
\bottomrule
\end{tabular}
\end{table}

### 2019-12-30 manuscript/main.Rmd
appendix changes

  row.names(change_stats) <- c(
   "Forest",
   "  -  of which edges",
   "Regrowth > 1958 loss",
   "Loss > 2000",
   "Loss > 1958 (permanent)")

```

@@

We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands, fields and buildings. We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e. forest or non-forested). This methodology is increasingly being used to automate pixel-level classification in (color) digital photography data [@buscombe2018]. Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from homogeneous forested or non-forested areas within the historical orthomosaic. Homogeneous tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3). We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing dataset. Source tiles did not repeat across datasets to limit overfitting. In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions. We applied this technique to 10% of the generated synthetic landscapes. The CNN model was trained for 100 epochs on a graphics processing unit (GPU) maximizing the Intersect-over-Union (IoU) using additional data augmentation. Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring. During final model evaluation we report the IoU of our out-of-sample test datasets.  The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote across overlapping areas to limit segmentation edge effects. We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow.

More details on the optimizer and other metrics etc.

@@

To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 data (GFC, tile 10N-020E) [@hansen2013]. Using this data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included forested pixels which recorded no loss throughout the whole 2000 - 2018 period. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30m GFC resolution. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. Where the datasets overlap we provide summary statistics of deforestation, reforestation and afforestation.

@@

We used an area of 400x400m (16 ha) around contemporary permanent sampling plot locations to quantify the general state of the canopy in both historical and contemporary remote sensing data. Contemporary data were resampled to the resolution of the historical data using a nearest neighbour interpolation for comparison. We used the first acquisition of a pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time). For permanent sampling plot locations in both historical and contemporary data a rectangular buffer of 200m around the centre of all plots was used to create small image subsets. All image subsets were processed using the R based FOTO implementation and package [@hufkens2019a] using global normalization, ensuring intercomparability of texture metrics across time and space [@Barbier2010]. We used a FOTO (moving) window size of 187 pixels (or ~150m), as described in @solorzano2018 and @Barbier2010, to ensure that multiple individual canopies could be included in the analysis. A buffer with a radius of 50m around the center of all permanent sampling plots was used to determine the average value of the first principle component (PC), explaining the majority of the variance in canopy texture. Where both the Geo-Eye data and the orthomosaic image overlapped we processed the intersecting region to explore a wall-to-wall comparison between past and current canopy texture metrics.

@@

Our CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in out-of-sample test data. Scaling our classifier to the whole historical orthomosaic we detected 162 km^2^ (or ~20% of the scene) of disturbed forests. A large fraction of the disturbed area was restored in the period between the two acquisitions. In total, 99 km^2^, or little over half of the affected forest was restored (Figure 5C-D, dark blue). Recent deforested areas, as registered through satellite remote sensing, approximate 88 km^2^ (Table1, Figure 5C-D / light green).

>>>>

  row.names(change_stats) <- c(
   "Forest",
   "  -  of which edges",
   "Regrowth > 1958 loss",
   "Loss > 2000",
   "Loss > 1958 (permanent)")
```

@@

We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands, fields and buildings. We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 pre-processing backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e. forest or non-forested). This methodology is increasingly being used to automate pixel-level classification in (color) digital photography data [@buscombe2018]. Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from homogeneous forested or non-forested areas within the historical orthomosaic. Homogeneous tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3). We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing dataset. Source tiles did not repeat across datasets to limit overfitting. In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions. We applied this technique to 10% of the generated synthetic landscapes. The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice and categorical cross-entropy loss functions. Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring. During final model evaluation we report the IoU of our out-of-sample test datasets. The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote across overlapping areas to limit segmentation edge effects. We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow. In addition, we used the first acquisition of a recent pan-chromatic Geo-Eye 1 stereo pair (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time) to classify and visually assess the robustness of the algorithm on contemporary remote sensing imagery with the same moving window approach.

@@

To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 data (GFC, tile 10N-020E) [@hansen2013]. Using this data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included forested pixels which recorded no loss throughout the whole 2000 - 2018 period. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30m GFC resolution. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. Where the datasets overlap we provide summary statistics of deforestation, reforestation and afforestation. The same methodology was used to quantify CNN model performance across the Geo-Eye 1 scene.

@@

We used an area of 400x400m (16 ha) around contemporary permanent sampling plot locations to quantify the general state of the canopy in both historical and contemporary pan-chromatic Geo-Eye 1 remote sensing data (GeoEye, 2012, Thornton, Colorado, U.S.A., order 737537, 2011-11-11 8:55h GMT or 9:55h local time). Geo-Eye data were resampled to the resolution of the historical data using a nearest neighbour interpolation for comparison. For permanent sampling plot locations in both historical and contemporary data a rectangular buffer of 200m around the centre of all plots was used to create small image subsets. All image subsets were processed using the R based FOTO implementation and package [@hufkens2019a] using global normalization, ensuring intercomparability of texture metrics across time and space [@Barbier2010]. We used a FOTO (moving) window size of 187 pixels (or ~150m), as described in @solorzano2018 and @Barbier2010, to ensure that multiple individual canopies could be included in the analysis. A buffer with a radius of 50m around the center of all permanent sampling plots was used to determine the average value of the first principle component (PC), explaining the majority of the variance in canopy texture. Where both the Geo-Eye data and the orthomosaic image overlapped we processed the intersecting region to explore a wall-to-wall comparison between past and current canopy texture metrics.

@@

Our CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in out-of-sample test data. A visual comparison on recent pan-chromatic Geo-Eye 1 data shows good agreement between the landsat based GFC data and downscaled CNN results (Appendix Figure 7).

[NOTE STATS]

Scaling our classifier to the whole historical orthomosaic we detected 162 km^2^ (or ~20% of the scene) of disturbed forests. A large fraction of the disturbed area was restored in the period between the two acquisitions. In total, 99 km^2^, or little over half of the affected forest was restored (Figure 5C-D, dark blue). Recent deforested areas, as registered through satellite remote sensing, approximate 88 km^2^ (Table1, Figure 5C-D / light green).
